---
title: "Baseball Report"
author: "Kyle Gilbert, Cole _, Dylan _"
highlight-style: github
format:
  html:
    code-fold: true
  pdf:
    geometry: 
      - top=30mm
      - left=30mm
editor: visual
---

```{r}
#| label: load-pkgs
#| code-summary: "Packages"
#| message: false
#| echo: false

library(tidyverse)
library(knitr)
library(ggplot2)
```

```{r}
#| label: load-datasets
#| echo: false

```

**Project Description:**

The goal of this final R project is to analyze the percentage of a team's salary cap allocated to specific players from 2000 to now and to explore the factors driving these players' compensation levels.

**Questions to answer:**

-   **Player Salary Insights**

    -   How does the percentage of salary cap each player earns correlate with their performance metrics (e.g., batting average, home runs, WAR)?

    -   What is the distribution of salaries across different positions (pitcher, catcher, etc.) for the Yankees compared to other teams?

-   **Team-Level Analysis**

    -   How does the Yankees' average player salary compare to that of other MLB teams?

    -   What percentage of the Yankees' salary cap is spent on top-performing players compared to the league average?

-   **Historical Trends**

    -   How has the percentage of the salary cap allocated to top-paid players changed over the years for the Yankees and other teams?

    -   Are there trends in salary allocation related to team performance in playoffs or championships?

-   **Position-Based Analysis**

    -   Are certain positions (e.g., pitchers, shortstops) allocated a higher percentage of the salary cap across MLB teams? How does this compare with the Yankees?

    -   Do players in specific positions receive more or less than their performance justifies?

-   **Team Strategy Insights**

    -   Does a higher percentage of salary cap spent on star players correlate with better overall team performance?

    -   How do the Yankees' salary allocation strategies compare to teams with similar or better performance records?

-   **League-Wide Comparisons**

    -   How do the Yankees rank in terms of salary cap utilization efficiency compared to other MLB teams?

    -   Are there any anomalies or outliers in salary allocations across the league?

-   **Contract Lengths and Salaries**

    -   Do longer contracts for Yankees players correlate with a higher percentage of the salary cap allocation?

    -   How does contract length influence the perceived value of players?

-   **Demographics and Salaries**

    -   Are there correlations between player age, experience, or nationality and the percentage of the salary cap they receive?

    -   How do rookie salaries as a percentage of salary caps compare between the Yankees and other teams?

-   **Revenue vs. Salaries**

    -   Does the Yankees' revenue have a stronger correlation with salary cap allocations compared to other teams?

    -   How does team revenue impact the overall spending strategy on players?
